Verification of Stencil Computations with Spatial Specifications

### 2016-07-07 paper.tex
dispersing the 'context' section

\usepackage[draft]{hyperref}
\hypersetup{
    unicode=false,          % non-Latin characters in Acrobat's bookmarks
    pdftoolbar=true,        % show Acrobat toolbar?
    pdfmenubar=true,        % show Acrobat menu?
    pdffitwindow=false,      % page fit to window when opened
    pdftitle={},    % title
    pdfauthor={}
    pdfsubject={},   % subject of the document
    pdfnewwindow=true,      % links in new window
    pdfkeywords={keywords}, % list of keywords
    colorlinks=true,       % false: boxed links; true: colored links
    linkcolor=darkblue,          % color of internal links
    citecolor=darkblue,        % color of links to bibliography
    filecolor=green,      % color of file links
    urlcolor=blue,          % color of external links
}

@@

\item we give a quantiative study of
  stencil computations in a corpus of numerical Fortran code,
  totalling one million lines of code. Our tool identifies and
  synthesises specifications for 35,000 stencil computations in the
  corpus. Approximately 10\% of the stencils we found are non-trivial,
  corresponding to code which is a possible source of errors;

\item we give a more detailed verification case study of two
  particular stencil computation and show that we
  are able to detect the vast majority of (synthesised) possible
  indexing errors.
\end{itemize}
%

@@

\paragraph{Context}

In recent years, there has been an increased awareness
of the need for formal verification techniques in
science~\cite{post2005computational,oberkampf2010verification,orchard2014computational}.
Our specification language is an approach in this direction, providing
a system of lightweight specification for numerical code.
The aim is to not only provide a way of exposing errors, but also
for specifying program behaviour to future-proof the code and ease
its maintenance. In order to increase the likelihood of impact, we aim
to integrate with existing practices smoothly, hence our
specifications are integrated as comments and our tool can synthesis
specifications for the user.

This paper considers the technical aspects of the language,
its semantics, inference, checking, and synthesis procedures.
We also evaluate our implementation on real scientific code 
used actively in research, ranging from small ($\sim$700loc) to large
($\sim$600kloc). 
 %Our case studies include the Unified Weather 
%model used by the UK Met Office~\cite{um-website} ($\approx{}600\textit{kloc}$) and other
%research-based numerical models, ranging from small to large.

@@

\begin{figure}[t]
\vspace{-0.1em}
\begin{align*}
\def\arraystretch{1.2}
\setlength{\arraycolsep}{0.2em}
\newcommand{\dimTy}{\mathbb{N}_{>0}}
\begin{array}{rl}
\nonterm{specification} ::= & \nonterm{regionDec} \mid \nonterm{specDec} \\
\nonterm{specDec} ::= & \term{stencil} \; \nonterm{spec} \;
                        \texttt{::} \; v \\
\nonterm{regionDec} ::= &  \texttt{region} \; \nonterm{rvar} \; \texttt{=} \;
                         \nonterm{region}\\[0.4em]
%\nonterm{spec} ::= & \nonterm{spatial} \mid \nonterm{temporal}
%\\[1em]
\nonterm{spec} ::= & [\nonterm{approx},] \; [\nonterm{mult},] \; \nonterm{region} \\
\nonterm{mult} ::= & \term{readOnce} \\
\nonterm{approx} ::= & \term{atMost} \; \mid \; \term{atLeast} \\[0.1em]
\nonterm{region} ::= & \nonterm{rvar} \; \\
\multicolumn{2}{l}{\qquad\qquad \mid \; \stenRefl{\dimTy}} \\
\multicolumn{2}{l}{\qquad\qquad \mid \;\stenFwd{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{irreflexive}]}} \\
\multicolumn{2}{l}{\qquad\qquad \mid \; \stenBwd{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{irreflexive}]}} \\
\multicolumn{2}{l}{\qquad\qquad \mid \; \stenCen{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{irreflexive}]}} \\
\multicolumn{2}{l}{\qquad\qquad \mid \; \nonterm{region} \; \term{+}
  \; \nonterm{region} \; \mid \; \nonterm{region} \; \term{*} \; \nonterm{region}} \\[0.5em]
%\multicolumn{2}{l}{\qquad\qquad \mid \; \nonterm{rvar}}  
%\\[0.5em]
%\nonterm{temporal} ::= \; & \term{dependency} \; (v \; \{ , v \}) [, \texttt{mutual}]
%  \\[0.5em]
\nonterm{rvar} ::= \; & [\text{\term{a}-\term{z}$\,$\term{A}-\term{Z}$\,$\term{0}-\term{9}}]+\\[-0.5em]
\end{array}
\end{align*}
\caption{Specification syntax (EBNF grammar)}
\label{fig:syntax}
\vspace{-0.8em}
\end{figure}

@@

\begin{figure}
%\vspace{-0.2em}
\begin{align*}
\setlength{\arraycolsep}{0.05em}
\begin{array}{c}
%\framebox{$\textit{region} \equiv \textit{region}'$} \\[0.6em]
\begin{array}{rll}
(\textsc{F\, \texttt{+} \,F}) \;\; &
\stenFwdSR{n \, \textsf{max} \, m}{d}{r \vee s} \\
 \equiv \; & \stenFwdSR{n}{d}{r} \; \texttt{+} \; \stenFwdSR{m}{d}{s} \\[0.6em]
%
%
(\textsc{B\, \texttt{+} \,B}) \;\; &
\stenBwdSR{n \, \textsf{max} \, m}{d}{r \vee s} \\
 \equiv \; & \stenBwdSR{n}{d}{r} \; \texttt{+} \; \stenBwdSR{m}{d}{s} \\[0.6em]
%
%
(\textsc{C\, \texttt{+} \,C}) \;\; &
\stenCenSR{n \, \textsf{max} \, m}{d}{r \vee s} \\
\equiv \; & \stenCenSR{n}{d}{r} \; \texttt{+} \; \stenCenSR{m}{d}{s} \\[0.6em]
%
%
(\textsc{C\, \texttt{+} \,F}) \;\; & \stenCenSR{n}{d}{r \vee s} \\
\textit{$m \leq n$} \; \equiv \; & \stenCenSR{n}{d}{r} \; \texttt{+} \;
                      \stenFwdSR{m}{d}{s} \\[0.6em]
%
%
(\textsc{C\, \texttt{+} \,B}) \;\; &
\stenCenSR{n}{d}{r \vee s} \\
\textit{$m \leq n$} \; \equiv \; & \stenCenSR{n}{d}{r} \; \texttt{+} \;
                      \stenBwdSR{m}{d}{s} \\[0.6em]
%
%
(\textsc{B\, \texttt{+} \,F}) \;\; &
\stenCenSR{n}{d}{r \vee s} \\
\equiv \; & \stenFwdSR{n}{d}{r} \; \texttt{+} \; \stenBwdSR{n}{d}{s}
  \\[0.6em]
%
%
(\textsc{R\, \texttt{+} \,F}) \;\; &
\stenFwdS{n}{d} \\
\equiv \; & \stenReflSR{d} \; \texttt{+} \; \stenFwdSR{n}{d}{r} \\[0.6em]
%
%
(\textsc{R\, \texttt{+} \,B}) \;\; &
\stenBwdS{n}{d} \\
\equiv \; & \stenReflSR{d} \; \texttt{+} \; \stenBwdSR{n}{d}{r} \\[0.6em]
%
%
(\textsc{R\, \texttt{+} \,C}) \;\; &
\stenCenS{n}{d} \\
\equiv \; & \stenReflSR{d} \; \texttt{+} \; \stenCenSR{n}{d}{r} \\[0.6em]
%
%
(\textsc{irrefl}) \;\; & (R \, \term{*} \, S^\irreflS) \, \term{+} \,
                         (R^\irreflS \, \term{*} \, S) \;
\equiv \; R \, \term{*} \, S \\[0.4em]
%
%
(\texttt{+}\textsc{idem}) \;\; & S \; \texttt{+} \; S \; \equiv \; S \\[0.4em]
%
(\texttt{+}\textsc{comm}) \;\; & S \; \texttt{+} \; T \; \equiv \; T \;
                       \texttt{+} \; S \\[0.4em]
%
(\texttt{+}\textsc{assoc}) \;\; & R \; \texttt{+} \; (S \; \texttt{+} \; T) \; \equiv \; (R \;
                       \texttt{+} \; S) \; \texttt{+} \; T \\[0.4em]
(\texttt{*}\textsc{comm}) \;\; & S \; \texttt{*} \; T \; \equiv \; T \;
                       \texttt{*} \; S \\[0.4em]
%
(\texttt{*}\textsc{assoc}) \;\; & R \; \texttt{*} \; (S \; \texttt{*} \; T) \; \equiv \; (R \;
                       \texttt{*} \; S) \; \texttt{*} \; T \\[0.4em]
(\textsc{dist}) \;\; & R \; \texttt{*} \; (S \; \texttt{+} \; T) \; \equiv \; (R \;
                       \texttt{*} \; S) \; \texttt{+} \; (R
                       \; \texttt{*} \; T)
\end{array}
%
%\framebox{${\textit{specDec} :: \texttt{v}} \equiv
%{\textit{specDec'} :: \texttt{v}}$} \\[0.7em]
%\begin{array}{rl}
% MUTUAL
%(\textsc{mutual}) \; &
%\texttt{stencil} \; \texttt{dependency} (\texttt{v}), \texttt{mutual}
%  :: \texttt{w}
%\\
%\equiv \; & \texttt{stencil} \; \texttt{dependency} (\texttt{w}), \texttt{mutual} ::
%  \texttt{v} \\[0.5em]
%(\textsc{coalesce}) \; &
%\texttt{stencil} \; \texttt{dependency} (\bar{v})
%  :: \texttt{v} \\
%& ; \texttt{stencil} \; \texttt{dependency} (\bar{w})
%  :: \texttt{v}
%\\
%\equiv \; & \texttt{stencil} \; \texttt{dependency} (\bar{v}, \bar{w}) ::
%  \texttt{v} \\[0.5em]
%\end{array}
\end{array}
\end{align*}
\caption{Equations on specifications (\emph{regions})}
\vspace{-1em}
\label{fig:equations}
\end{figure}

Figure~\ref{fig:equations} lists the equational theory for
specifications which is defined by the binary relation $\equiv$
between region terms. The equations use abbreviations
\term{refl}, \term{fwd}, \term{bwd}, \term{cen}, and \irreflS{}
for \term{reflexive}, \term{forward}, \term{backward},
\term{centered}, and \irrefl{} respectively. Furthermore, the abbreviations include syntactic sugar
for quantifying over the absence or presence of the \irrefl{}
attribute on region constants as a boolean. For example,
the syntactic sugar for \term{fwd} is defined:
%%
\begin{align*}
\hspace{-0.25em}
& \stenFwdSR{n}{d}{r}
\; := \; \\[-0.4em]
& \qquad \begin{cases}
\stenFwd{n}{d}{, \irrefl} & \;\;\textit{if} \;\, \neg \, r \\
\stenFwd{n}{d}{}  & \;\;\textit{if} \;\,  r
\end{cases} \\
\textit{or} \;\;
& \stenFwdS{n}{d} := \stenFwd{n}{d}{}
\end{align*}
%
and similarly for \term{bwd} and \term{cen}.

Stencil specifications in \textit{spec} are considered
equal when they have the same modifiers and $\equiv$ equates
their regions, \ie{}:
\begin{align*}
(\textit{region} \equiv \textit{region'}) 
 \Rightarrow \quad\;\; & \texttt{stencil} \; [\textit{approx},] \; [\textit{mult},] \;
\textit{region} \\[-0.4em]
\quad \equiv \; & \texttt{stencil} \; [\textit{approx},] \;
            [\textit{mult},] \; \textit{region}'
\end{align*}

The equations reveal a part of the semantics for our
specification language, which we briefly unpack.

@@

\begin{figure}[t]
\begin{align*}
\hspace{-0.5em}
\begin{array}{c}
%%
%% INEQUATIONS ON REGION
%% 
%\framebox{$\textit{region} <: \textit{region}'$} \\[0.85em]
%
%\dfrac{}{R <:^r R}\trule{\textsc{refl}} \qquad \dfrac{P <:^r R \quad R <:^r Q}{P <:^r
%  Q}\trule{\textsc{trans}} 
%\qquad
%\\[1.5em]
\vspace{-0.4em}
\setlength{\arraycolsep}{0.1em}
\dfrac{\hspace{3em} m \leq n \qquad\qquad r \Rightarrow s \hspace{3em}}
{\begin{array}{rl}
\stenFwdSR{m}{d}{r} & <:^r \stenFwdSR{n}{d}{s}
%& \;\, ds \subseteq es \wedge n \leq m
\\
\wedge \; \stenBwdSR{m}{d}{r} & <:^r \stenBwdSR{n}{d}{s}
%& \;\, ds \subseteq es \wedge n \leq m
\\
\wedge \; \stenCenSR{m}{d}{r} & <:^r \stenCenSR{n}{d}{s}
%& \;\, ds \subseteq es \wedge n \leq m
\\
\wedge \; \stenBwdSR{m}{d}{r} & <:^r \stenCenSR{n}{d}{s} \\
\wedge \; \stenFwdSR{m}{d}{r} & <:^r \stenCenSR{n}{d}{s} \\[1.2em]
\end{array}}\trule{\textsc{over}} \\
\hspace{-0.5em}
\dfrac{P_1 <:^r R_1 \quad P_2 <:^r R_2}
      {P_1 \, \texttt{+} \, P_2 <:^r R_1 \, \texttt{+} \, R_2}
\trule{\textsc{cong}\texttt{+}} \;\;\;
\dfrac{P_1 <:^r R_1 \quad P_2 <:^r R_2}
      {P_1 \, \texttt{*} \, P_2 <:^r R_1 \, \texttt{*} \, R_2}
\trule{\textsc{cong}\texttt{*}} \\[1.5em]
%%
%% INEQUATIONS ON SPEC
%%
%\framebox{$\textit{spec} <: \textit{spec}'$} \\[0.8em]
%
\dfrac{}{\term{readOnce} \, R <: R}\trule{\textsc{rep}}
\quad
\dfrac{}{\term{atLeast} \, R <: \term{atMost} \,
  R}\trule{\textsc{bound}} \\[1.5em]
%%
\dfrac{P <:^r R}{\term{atLeast}  P <: \term{atLeast} 
  R}\trule{\textsc{cong}\texttt{L}}
\ 
\dfrac{P <:^r R}{\term{atMost}  P <: \term{atMost} 
  R}\trule{\textsc{cong}\texttt{M}} \\[1.5em]
\dfrac{P <:^r R}{\term{readOnce} \, P <: \term{readOnce} \,
  R}\trule{\textsc{cong}\texttt{R}} 
\quad
\dfrac{R \equiv S}{R <:^r S}\trule{\textsc{eq}} 
%%
\end{array}
\end{align*}
\caption{Inequations $<:^r$ on \textit{regions} and
$<:$ on \textit{specs} (omitting reflexivity
and transitivity). }
\label{fig:inequations}
\vspace{-0.8em}
\end{figure}

@@

%\newcommand{\mkBig}[2]{\overset{\LARGE{\term{#1}}}{#2}}
\begin{definition} The $\odot$ operation on $\textsf{A}(\specDNF{})$
  is defined in terms of the intermediate operator $\hat{\odot}$ on
  \specDNF{}, defined:
%%
\begin{align*}
S \,\hat{\odot}\, \epsilon = S \quad
\epsilon \,\hat{\odot}\, S = S \qquad
S \,\hat{\odot}\, T = \bplus_{(\tyProd{s}, \tyProd{t}) \in (S \times
   T)} (\tyProd{s} \; \term{*} \; \tyProd{t})
\end{align*}
%%
That is, the $\hat{\odot}$ product of two specifications $S$ and $T$ in DNF form 
is the $\term{+}$ sum of all pairwise $\term{*}$ products for every
pair of $\tyProd{spec}$ drawn from $S$ and $T$. The result
is indeed in DNF-form, where $\hat{\odot}$ is essentially applying
 the (\trule{\textsc{dist}}) rule of our equational theory
 (Fig.~\ref{fig:equations},p.~\pageref{fig:equations}). The operation is
commutative and associative, and sound with respect to our
model (see Lemma~\ref{lem:alg-soundness} below).
The $\hat{\odot}$ operation is then lifted to all combinations of
\textsf{Approx} for $\odot$:
\begin{align*}
\textsf{lower}(S) \odot \textsf{exact}(T) = \; &
       %\textsf{exact}(T) \odot \textsf{lower}(S) = \;
       \textsf{both} \, (S \hat{\odot} T, \; T) \\
\textsf{upper}(S) \odot \textsf{exact}(T) = \; & 
       %\textsf{exact}(T) \odot \textsf{upper}(S) =
       \textsf{both} \, (T, \; S \hat{\odot} T) \\
\textsf{lower}(S) \odot \textsf{both}(T_l,T_u) = \; &
        %\textsf{both}(T_l,T_u) \odot \textsf{lower}(S) = \;
        \textsf{both} \, (S \hat{\odot} T_l, \; T_u) \\
\textsf{upper}(S) \odot \textsf{both}(T_l,T_u) = \; & \textsf{both} \,
                                                      (T_l, \; S \hat{\odot} T_u) \\
\textsf{both}(S_l,S_u) \odot
\textsf{both}(T_l,T_u) & = \textsf{both} (S_l \hat{\odot} T_l, \; S_u
                         \hat{\odot} T_u) \\
\textit{inj}(S) \odot \textit{inj}(T) = \; & \textit{inj}(S
                                              \hat{\odot} T)
\end{align*}
where in the last case $\textit{inj}$ corresponds to the unary
injections of \textsf{Approx}. For brevity we omit the cases where the parameters above are flipped
as $\odot$ is commutative.
\end{definition}

@@

We presented in this work an specification language 
for the access pattern of a stencil computation. Compared to the
related work (discussed above), our language is novel in that it does
not aim to replicate indexing in the specification, but instead
provides abstract spatial combinators which capture a large number
of common patterns.

>>>>

\usepackage{hyperref}
\hypersetup{
    unicode=false,          % non-Latin characters in Acrobat's bookmarks
    pdftoolbar=true,        % show Acrobat toolbar?
    pdfmenubar=true,        % show Acrobat menu?
    pdffitwindow=false,      % page fit to window when opened
    pdftitle={},    % title
    pdfauthor={}
    pdfsubject={},   % subject of the document
    pdfnewwindow=true,      % links in new window
    pdfkeywords={keywords}, % list of keywords
    colorlinks=true,       % false: boxed links; true: colored links
    linkcolor=darkblue,          % color of internal links
    citecolor=darkblue,        % color of links to bibliography
    filecolor=green,      % color of file links
    urlcolor=blue,          % color of external links
}

@@

\item we perform a quantitative study of
  stencil computations in a corpus of numerical Fortran programs
  (ranging from small to large),  totalling one million lines of code. Our tool identifies and
  synthesises specifications for 35,000 stencil computations in the
  corpus. Approximately 10\% of the stencils we found are non-trivial,
  corresponding to code which is a possible source of errors;

\item we give a detailed verification case study of two
  particular stencil computation, simulating programming errors.
  Our approach is able to detect the vast majority of possible
  indexing errors.
\end{itemize}
%

@@

%This paper considers the technical aspects of the language,
%its semantics, inference, checking, and synthesis procedures.
%We also evaluate our implementation on real scientific code 
%used actively in research, ranging from small ($\sim$700loc) to large
%($\sim$600kloc). 
 %Our case studies include the Unified Weather 
%model used by the UK Met Office~\cite{um-website} ($\approx{}600\textit{kloc}$) and other
%research-based numerical models, ranging from small to large.

@@

\begin{figure}[t]
\vspace{-0.5em}
\begin{align*}
\def\arraystretch{1.2}
\setlength{\arraycolsep}{0.2em}
\newcommand{\dimTy}{\mathbb{N}_{>0}}
\begin{array}{rl}
\nonterm{specification} ::= & \nonterm{regionDec} \mid \nonterm{specDec} \\
\nonterm{specDec} ::= & \term{stencil} \; \nonterm{spec} \;
                        \texttt{::} \; v \\
\nonterm{regionDec} ::= &  \texttt{region} \; \nonterm{rvar} \; \texttt{=} \;
                         \nonterm{region}\\[0.4em]
%\nonterm{spec} ::= & \nonterm{spatial} \mid \nonterm{temporal}
%\\[1em]
\nonterm{spec} ::= & [\nonterm{approx},] \; [\nonterm{mult},] \; \nonterm{region} \\
\nonterm{mult} ::= & \term{readOnce} \\
\nonterm{approx} ::= & \term{atMost} \; \mid \; \term{atLeast} \\[0.1em]
\nonterm{region} ::= & \nonterm{rvar} \; \\
\multicolumn{2}{l}{\qquad\qquad \mid \; \stenRefl{\dimTy}} \\
\multicolumn{2}{l}{\qquad\qquad \mid \;\stenFwd{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{irreflexive}]}} \\
\multicolumn{2}{l}{\qquad\qquad \mid \; \stenBwd{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{irreflexive}]}} \\
\multicolumn{2}{l}{\qquad\qquad \mid \; \stenCen{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{irreflexive}]}} \\
\multicolumn{2}{l}{\qquad\qquad \mid \; \nonterm{region} \; \term{+}
  \; \nonterm{region} \; \mid \; \nonterm{region} \; \term{*} \; \nonterm{region}} \\[0.5em]
%\multicolumn{2}{l}{\qquad\qquad \mid \; \nonterm{rvar}}  
%\\[0.5em]
%\nonterm{temporal} ::= \; & \term{dependency} \; (v \; \{ , v \}) [, \texttt{mutual}]
%  \\[0.5em]
\nonterm{rvar} ::= \; & [\text{\term{a}-\term{z}$\,$\term{A}-\term{Z}$\,$\term{0}-\term{9}}]+\\[-0.5em]
\end{array}
\end{align*}
\caption{Specification syntax (EBNF grammar)}
\label{fig:syntax}
\vspace{-0.8em}
\end{figure}

@@

\begin{figure}
%\vspace{-0.2em}
\begin{align*}
\setlength{\arraycolsep}{0.05em}
\begin{array}{c}
%\framebox{$\textit{region} \equiv \textit{region}'$} \\[0.6em]
\begin{array}{rll}
(\textsc{F\, \texttt{+} \,F}) \;\; &
\stenFwdSR{n \, \textsf{max} \, m}{d}{r \vee s} \\
 \equiv \; & \stenFwdSR{n}{d}{r} \; \texttt{+} \; \stenFwdSR{m}{d}{s} \\[0.6em]
%
%
(\textsc{B\, \texttt{+} \,B}) \;\; &
\stenBwdSR{n \, \textsf{max} \, m}{d}{r \vee s} \\
 \equiv \; & \stenBwdSR{n}{d}{r} \; \texttt{+} \; \stenBwdSR{m}{d}{s} \\[0.6em]
%
%
(\textsc{C\, \texttt{+} \,C}) \;\; &
\stenCenSR{n \, \textsf{max} \, m}{d}{r \vee s} \\
\equiv \; & \stenCenSR{n}{d}{r} \; \texttt{+} \; \stenCenSR{m}{d}{s} \\[0.6em]
%
%
(\textsc{C\, \texttt{+} \,F}) \;\; & \stenCenSR{n}{d}{r \vee s} \\
\textit{$m \leq n$} \; \equiv \; & \stenCenSR{n}{d}{r} \; \texttt{+} \;
                      \stenFwdSR{m}{d}{s} \\[0.6em]
%
%
(\textsc{C\, \texttt{+} \,B}) \;\; &
\stenCenSR{n}{d}{r \vee s} \\
\textit{$m \leq n$} \; \equiv \; & \stenCenSR{n}{d}{r} \; \texttt{+} \;
                      \stenBwdSR{m}{d}{s} \\[0.6em]
%
%
(\textsc{B\, \texttt{+} \,F}) \;\; &
\stenCenSR{n}{d}{r \vee s} \\
\equiv \; & \stenFwdSR{n}{d}{r} \; \texttt{+} \; \stenBwdSR{n}{d}{s}
  \\[0.6em]
%
%
(\textsc{R\, \texttt{+} \,F}) \;\; &
\stenFwdS{n}{d} \\
\equiv \; & \stenReflSR{d} \; \texttt{+} \; \stenFwdSR{n}{d}{r} \\[0.6em]
%
%
(\textsc{R\, \texttt{+} \,B}) \;\; &
\stenBwdS{n}{d} \\
\equiv \; & \stenReflSR{d} \; \texttt{+} \; \stenBwdSR{n}{d}{r} \\[0.6em]
%
%
(\textsc{R\, \texttt{+} \,C}) \;\; &
\stenCenS{n}{d} \\
\equiv \; & \stenReflSR{d} \; \texttt{+} \; \stenCenSR{n}{d}{r} \\[0.6em]
%
%
(\textsc{irrefl}) \;\; & (R \, \term{*} \, S^\irreflS) \, \term{+} \,
                         (R^\irreflS \, \term{*} \, S) \;
\equiv \; R \, \term{*} \, S \\[0.4em]
%
%
(\texttt{+}\textsc{idem}) \;\; & S \; \texttt{+} \; S \; \equiv \; S \quad
%
(\texttt{+}\textsc{comm}) \;\; S \; \texttt{+} \; T \; \equiv \; T \;
                       \texttt{+} \; S \\[0.4em]
%
(\texttt{+}\textsc{assoc}) \;\; & R \; \texttt{+} \; (S \; \texttt{+} \; T) \; \equiv \; (R \;
                       \texttt{+} \; S) \; \texttt{+} \; T \\[0.4em]
(\texttt{*}\textsc{comm}) \;\; & S \; \texttt{*} \; T \; \equiv \; T \;
                       \texttt{*} \; S \\[0.4em]
%
(\texttt{*}\textsc{assoc}) \;\; & R \; \texttt{*} \; (S \; \texttt{*} \; T) \; \equiv \; (R \;
                       \texttt{*} \; S) \; \texttt{*} \; T \\[0.4em]
(\textsc{dist}) \;\; & R \; \texttt{*} \; (S \; \texttt{+} \; T) \; \equiv \; (R \;
                       \texttt{*} \; S) \; \texttt{+} \; (R
                       \; \texttt{*} \; T)
\end{array}
%
%\framebox{${\textit{specDec} :: \texttt{v}} \equiv
%{\textit{specDec'} :: \texttt{v}}$} \\[0.7em]
%\begin{array}{rl}
% MUTUAL
%(\textsc{mutual}) \; &
%\texttt{stencil} \; \texttt{dependency} (\texttt{v}), \texttt{mutual}
%  :: \texttt{w}
%\\
%\equiv \; & \texttt{stencil} \; \texttt{dependency} (\texttt{w}), \texttt{mutual} ::
%  \texttt{v} \\[0.5em]
%(\textsc{coalesce}) \; &
%\texttt{stencil} \; \texttt{dependency} (\bar{v})
%  :: \texttt{v} \\
%& ; \texttt{stencil} \; \texttt{dependency} (\bar{w})
%  :: \texttt{v}
%\\
%\equiv \; & \texttt{stencil} \; \texttt{dependency} (\bar{v}, \bar{w}) ::
%  \texttt{v} \\[0.5em]
%\end{array}
\end{array}
\end{align*}
\caption{Equations on specifications (\emph{regions})}
\vspace{-1em}
\label{fig:equations}
\end{figure}

Figure~\ref{fig:equations} lists the equational theory for
specifications given by the binary relation $\equiv$
on \emph{region} terms. The equations use abbreviations
\term{refl}, \term{fwd}, \term{bwd}, \term{cen}, and \irreflS{}
for \term{reflexive}, \term{forward}, \term{backward},
\term{centered}, and \irrefl{} respectively. Furthermore, the abbreviations include syntactic sugar
for quantifying over the absence or presence of the \irrefl{}
attribute on region constants as a boolean. For example,
\term{fwd} syntactic sugar is defined:
%%
\begin{align*}
\hspace{-0.25em}
& \stenFwdSR{n}{d}{r}
\; := \; \\[-0.4em]
& \qquad \begin{cases}
\stenFwd{n}{d}{, \irrefl} & \;\;\textit{if} \;\, \neg \, r \\[-0.1em]
\stenFwd{n}{d}{}  & \;\;\textit{if} \;\,  r
\end{cases} \\
\textit{or} \;\;
& \stenFwdS{n}{d} := \stenFwd{n}{d}{}
\end{align*}
%
and similarly for \term{bwd} and \term{cen}.

Stencil specifications in \textit{spec} are considered
equal when they have the same modifiers and $\equiv$ equates
their regions, \ie{}:
\begin{align*}
(\textit{region} \equiv \textit{region'}) 
 \Rightarrow \quad\;\; & \texttt{stencil} \; [\textit{approx},] \; [\textit{mult},] \;
\textit{region} \\[-0.4em]
\quad \equiv \; & \texttt{stencil} \; [\textit{approx},] \;
            [\textit{mult},] \; \textit{region}'
\end{align*}
%
The equations reveal a part of the semantics for our
specification language, which we briefly unpack.

@@

\begin{figure}[t]
\begin{align*}
\hspace{-0.5em}
\begin{array}{c}
%%
%% INEQUATIONS ON REGION
%% 
%\framebox{$\textit{region} <: \textit{region}'$} \\[0.85em]
%
%\dfrac{}{R <:^r R}\trule{\textsc{refl}} \qquad \dfrac{P <:^r R \quad R <:^r Q}{P <:^r
%  Q}\trule{\textsc{trans}} 
%\qquad
%\\[1.5em]
\vspace{-0.4em}
\setlength{\arraycolsep}{0.1em}
\dfrac{\hspace{3em} m \leq n \qquad\qquad r \Rightarrow s \hspace{3em}}
{\begin{array}{rl}
\stenFwdSR{m}{d}{r} & <:^r \stenFwdSR{n}{d}{s}
%& \;\, ds \subseteq es \wedge n \leq m
\\
\wedge \; \stenBwdSR{m}{d}{r} & <:^r \stenBwdSR{n}{d}{s}
%& \;\, ds \subseteq es \wedge n \leq m
\\
\wedge \; \stenCenSR{m}{d}{r} & <:^r \stenCenSR{n}{d}{s}
%& \;\, ds \subseteq es \wedge n \leq m
\\
\wedge \; \stenBwdSR{m}{d}{r} & <:^r \stenCenSR{n}{d}{s} \\
\wedge \; \stenFwdSR{m}{d}{r} & <:^r \stenCenSR{n}{d}{s} \\[1.2em]
\end{array}}\trule{\textsc{over}} \\[-0.2em]
\hspace{-0.5em}
\dfrac{P_1 <:^r R_1 \quad P_2 <:^r R_2}
      {P_1 \, \texttt{+} \, P_2 <:^r R_1 \, \texttt{+} \, R_2}
\trule{\textsc{cong}\texttt{+}} \;\;\;
\dfrac{P_1 <:^r R_1 \quad P_2 <:^r R_2}
      {P_1 \, \texttt{*} \, P_2 <:^r R_1 \, \texttt{*} \, R_2}
\trule{\textsc{cong}\texttt{*}} \\[1.5em]
%%
%% INEQUATIONS ON SPEC
%%
%\framebox{$\textit{spec} <: \textit{spec}'$} \\[0.8em]
%
\dfrac{}{\term{readOnce} \, R <: R}\trule{\textsc{rep}}
\quad
\dfrac{}{\term{atLeast} \, R <: \term{atMost} \,
  R}\trule{\textsc{bound}} \\[1.5em]
%%
\dfrac{P <:^r R}{\term{atLeast}  P <: \term{atLeast} 
  R}\trule{\textsc{cong}\texttt{L}}
\ 
\dfrac{P <:^r R}{\term{atMost}  P <: \term{atMost} 
  R}\trule{\textsc{cong}\texttt{M}} \\[1.5em]
\dfrac{P <:^r R}{\term{readOnce} \, P <: \term{readOnce} \,
  R}\trule{\textsc{cong}\texttt{R}} 
\quad
\dfrac{R \equiv S}{R <:^r S}\trule{\textsc{eq}} 
%%
\end{array}
\end{align*}
\caption{Inequations $<:^r$ on \textit{regions} and
$<:$ on \textit{specs} (omitting reflexivity
and transitivity). }
\label{fig:inequations}
\vspace{-0.8em}
\end{figure}

@@

%\newcommand{\mkBig}[2]{\overset{\LARGE{\term{#1}}}{#2}}
\begin{definition} The $\odot$ operation on $\textsf{A}(\specDNF{})$
  is defined in terms of the intermediate operator $\hat{\odot}$ on
  \specDNF{}, defined:
%%
\begin{align*}
S \,\hat{\odot}\, \epsilon = S \quad
\epsilon \,\hat{\odot}\, S = S \qquad
S \,\hat{\odot}\, T = \bplus_{(\tyProd{s}, \tyProd{t}) \in (S \times
   T)} (\tyProd{s} \; \term{*} \; \tyProd{t})
\end{align*}
%%
That is, the $\hat{\odot}$ product of two specifications $S$ and $T$ in DNF form 
is the $\term{+}$ sum of all pairwise $\term{*}$ products for every
pair of $\tyProd{spec}$ drawn from $S$ and $T$. The result
is indeed in DNF-form, where $\hat{\odot}$ is essentially applying
 the (\trule{\textsc{dist}}) rule of our equational theory
 (Fig.~\ref{fig:equations},p.~\pageref{fig:equations}). The operation is
commutative and associative, and sound with respect to our
model (see Lemma~\ref{lem:alg-soundness} below).
The $\hat{\odot}$ operation is then lifted to all combinations of
\textsf{Approx} for $\odot$:
\begin{align*}
\textsf{lower}(S) \odot \textsf{exact}(T) = \; &
       %\textsf{exact}(T) \odot \textsf{lower}(S) = \;
       \textsf{both} \, (S \hat{\odot} T, \; T) \\
\textsf{upper}(S) \odot \textsf{exact}(T) = \; & 
       %\textsf{exact}(T) \odot \textsf{upper}(S) =
       \textsf{both} \, (T, \; S \hat{\odot} T) \\
\textsf{lower}(S) \odot \textsf{both}(T_l,T_u) = \; &
        %\textsf{both}(T_l,T_u) \odot \textsf{lower}(S) = \;
        \textsf{both} \, (S \hat{\odot} T_l, \; T_u) \\
\textsf{upper}(S) \odot \textsf{both}(T_l,T_u) = \; & \textsf{both} \,
                                                      (T_l, \; S \hat{\odot} T_u) \\
\textsf{both}(S_l,S_u) \odot
\textsf{both}(T_l,T_u)  = \; & \textsf{both} (S_l \hat{\odot} T_l, \; S_u
                         \hat{\odot} T_u) \\
\textit{inj}(S) \odot \textit{inj}(T) = \; & \textit{inj}(S
                                              \hat{\odot} T)
\end{align*}
where in the last case $\textit{inj}$ corresponds to the unary
injections of \textsf{Approx}. For brevity we omit the cases where the arguments above are flipped
as $\odot$ is commutative.
\end{definition}

@@

In recent years, there has been an increased awareness
of the need for formal verification techniques in
science~\cite{post2005computational,oberkampf2010verification,orchard2014computational}.
Our specification language is an approach in this direction, providing
a system of lightweight specification for numerical code.
Compared to other specification approaches (discussed above), our language is novel in that it does
not aim to replicate indexing in the specification, but instead
provides abstract spatial combinators which capture a large number
of common patterns.

Our aim is to not only provide a way of exposing errors, but also
for specifying program behaviour to future-proof the code and ease
its maintenance. To increase the likelihood of impact, we 
integrate smoothly with existing practices, providing 
specifications as comments and specification synthesis
specifications in the the tool.
